Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-12-11 paper/paper_acm.tex
add ACM version

\documentclass[conference]{IEEEtran}

@@

% *** MATH PACKAGES ***
%
\usepackage[cmex10]{amsmath}
\usepackage{amssymb,amsthm,amsfonts}

\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{defi}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{ax}[thm]{Axiom}
\newtheorem{cond}[thm]{Condition}

@@

 \IEEEpubid{}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Lawvere-Tierney Sheafification\\ in Homotopy Type Theory}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Kevin Quirin and Nicolas Tabareau}
\IEEEauthorblockA{Inria \\ \'Ecole des Mines de Nantes \\ Nantes, France}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}

  Sheafification is a popular tool in topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been developed in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher toposes.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, being
  compatible with the full type-theoretic axiom of choice.

\end{abstract}

% no keywords

@@

One of the most direct application of homotopical notion to type
theory is the introduction of homotopy types. 
%
Using the analogy that points in a space correspond to elements of a
type and that paths between two points correspond to 
elements of the corresponding identity type (which defines equality in type theory),  
%
an $n$-type is simply a type
for which equality becomes trivial up to level $n$. 
%
Voevodsky has realized that this notion admits a compact inductive definition
internal to type theory, given by
% Our work is mainly based on the stratification by $\Type_n$~:
\begin{defi}
  $\IsType n$ is defined by induction on $n\geqslant -2$:
  \begin{itemize}
  \item $\IsType {(-2)} X$ if $X$ is a contractible type, \ie $X$
    is pointed by $c:X$, and every other point in $X$ is connected to $c$.
  \item $\IsType {(n+1)} X = \prod_{x,y:X} \IsType n (x=y).$
  \end{itemize}
  Then, $\Type_n = \sum_{X:\Type} \IsType n X$.
\end{defi}
% For a type $X$, to be in $\Type_n$ means that path spaces of $X$ are
% trivial after $n+1$ iteration.
%
We also define some syntactic sugar for contractible types and mere propositions.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
$$
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
$$ 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
$$\chi : \left(\sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b}\right) \xrightarrow{\sim} 
 (B \to \Type_n)$$
%
 such that the diagram
$$
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
$$
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
$$t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).$$

@@

Then $P$ is the {\em colimit} of $D$ if for any other $X:\Type$, the
map
$$\lambda f:P \to X, \left( \lambda i,~f \circ q_i~;~ \lambda i\, j\,
  \phi,~ \precompose_f (p(i, j, g)) \right)$$
where $\precompose_f : \phi = \psi \to \phi \circ f = \psi \circ f$,
is an equivalence.
\end{defi}

@@

\begin{lem}
  Let $D$ be a diagram, $m$ a truncation index, and
  $P:\Type_{m}$. Let's call $P$ a $m$-colimit of $D$ if
  for any other $X:\Type_m$, the
  map
  $$\lambda f:P \to X, \left( \lambda i,~f \circ q_i~;~ \lambda i\, j\,
    \phi,~ \precompose_f (p(i, j, g)) \right)$$
  is an equivalence.

@@

\begin{defi}
  Let $f:X \to Y$ be a map. $T_f$ is the higher inductive type given
  by
  \begin{itemize}
  \item $t:~X \to T_f$
  \item $\alpha:~\forall a\,b:X,~f a = f b \to t a = t b$
  \item $\alpha_1:~\forall a:X,~\alpha\, a\, a\, 1 = 1$
  \end{itemize}
  We view $T_f$ as the coequalizer of
  $$\xymatrix{
    \sum_{a,b:X} f a = f b \dar[r]{4pt}^-{\pi_1}_-{\pi_2} & X
  }$$
  preserving the identity.

@@

Then, the desired diagram is the mapping telescope of the iterations
of $T$.
\begin{defi}
  Let $f$ be a map from $X$ to $Y$. Then $C(f)$
  is given by the diagram
  $$\xymatrix{
    C(f) := X \ar[r]^t & T_f \ar[r]^t & T_{\tilde f} \ar[r] &\cdots
  }$$
\end{defi}

@@

The proof is based on the following central result.
\begin{lem}
  Let $X:\Type_p$, $Y:\Type_p^\modal$ and $f:X\to Y$. If for all $y:Y$,
  $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
\end{lem}
%
It is straighforward to define a map
$$\phi:\sum_{x:X}  (f x = y)\to
\sum_{x:\modal X} (\modal f x = \eta_Y y),$$
using $\eta$ functions.
We just need to check that every $\modal(\fib \phi {x;p})$ is
contractible.
Technical transformations allow one to prove
$$\fib\phi{x;p} = \fib s {y;p^{-1}}$$
for some $a:\modal X$, $b:\modal Y$ and 
$$s:\fib{\eta_X}a \to \fib{\eta_Y}b.$$
%
From definition of left exactness, one can deduce the following:
\begin{lem}
  Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
  contractible, then so is $\fib f b$ for any $b:B$.
\end{lem}
Thus, we just need to prove that $\modal(\fib {\eta_X} a)$ and
$\modal(\fib {\eta_Y} b)$ are contractible.
But we have the lemma:
\begin{lem}
  Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
  y)$ is contractible.
\end{lem}

@@

Here, we need to distinguish between
dense $(-1)$-subobjects, that will be used in the definition of
sheaves, and dense $n$-subobjects, that will be used in the definition
of separated types.

@@

  We define $g$ as 
  $$g(e) = \modal_n \left( \fib \phi e \right), $$
  where
  $$\phi : \left(\sum_{b:\left(\sum_{e:E} (\chi\, e)\right)} (f\,
    b)\right) \to E$$
  defined by $\phi(x) = (x_1)_1$.
  Using the following properties on fibers, we can prove that it
  defines an equivalence.
  \begin{lem}
    Let $A,B,C:\Type_n$, $f:A\to B$ and $g:B\to C$.
    Then
    \begin{itemize}
    \item If $g$ is an embedding, then 
      $$\forall b:B,~\fib f b = \fib {g\circ f} {g\, b}.$$
    \item If all fibers of $f$ and $g$ are $n$-truncated, then
      $$\forall c:C,~\modal_n(\fib {g \circ f} c) =
      \modal_n \left(  
        \sum_{w:\fib g c} \modal_n (\fib f {w_1})
      \right).$$
    \end{itemize}
  \end{lem}

@@

Let $T : \Type_{n+1}$. We define $\separated T$ as the image (see
Section~\ref{sec:epi-mono-fact}) of
$\modal_n^T \circ \{\cdot\}_T$, as in
$$\xymatrix{
  T \ar[r]^{\{\cdot\}_T} \ar[d]_{\mu_T} & \Type_n^T \ar[d]^{\modal_n^T} \\
  \separated T \ar[r]& \left( \Type_n^\modal \right)^T
}, $$
where $\{\cdot\}_T$ is the singleton map $\lambda (t:T),~\lambda
(t':T),~t=t'$.
%
$\separated T$ can be given explicitly by
%
$$
\begin{array}{rcl}
\separated T &\defeq & \im (\lambda~t:T,~\lambda~ t',~ \modal_n (t = t')) \\
          & \defeq & \sum_{u:T \to \Type_n^\modal} \left\| \sum_{a:T} 
            (\lambda t,~\modal_n (a=t)) = u\right\|.
\end{array}
$$
%
This corresponds to the free separated object used in the topos-theoretic construction, but using $\Type_n^\modal$ instead of the
$j$-subobject classifier $\Omega_j$.
%
\begin{prop}
  For any $T:\Type_{n+1}$, $\separated T$ is separated.  
\end{prop}

@@

  We use the encode-decode method to characterize $\tilde f |a|_{n+1}
  = x$, and the result follows. We refer to the formalization for details.
\end{proof}

@@

\section{Extension to $\Type$}
\label{sec:extension-type}

Now, we have defined a (countably) infinite family of modalities
$\Type_i \to \Type_i$. Now:

\begin{lem}
  Let $\modal_i:\Type_i \to \Type_i$ be a modality. Then $\modal_i
  \circ \|\cdot\|_i : \Type \to \Type$ is a modality in the sense
  of~\cite[Section 7.7]{hottbook}
\end{lem}

Taking $i$ as a non-fixed parameter, that would allow us to work in an
universe where the forced principle ({\em e.g.} mere excluded middle) is
true for any explicit truncated type. Indeed, $i$ can be chosen
dynamically along a proof, and thus be increased as much as needed.

\subsection{New objects}
\label{sec:new-objects}

Working in $\Type^\modal$, one can consider new objects. For example,
let $c:nat \to \nat$ implementing the Collatz function. Then, for any
$n:\nat$, the type $C_n \defeq \sum_{k:\nat}\left( c^k(n) = 1 \land k\text{ is the
  smallest such $\nat$}\right)$ is an $\HProp$.
Thus, for any $n$, one have $C_n + \lnot C_n$, and one can {\em
  constructively} build a map $\varphi:\nat \to \two$ such that
$\varphi(n) = \True$ if and only if $n$ satisfies the Collatz
conjecture.
Actually, one can defined such a map for any family of $\HProp$
indexed by $\nat$, for example with the family $C_n \defeq \text{«\,The $n$-th Turing Machine halts\,»}$.

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification}.
%
It is based on the Coq/HoTT library available at
\url{https://github.com/HoTT/HoTT} and uses the version of Coq
tailored for that library.
%
After reviewing the content and some statistics about the
formalization in Section~\ref{sec:cont-form}, we present the
limitations of our formalization in Section~\ref{sec:limit-form}, in
particular the issues relative to universe polymorphism.

@@

We provide a more detailed insight of the structure of our formalization:
\begin{itemize} 
\item  Homotopy fibers and subobject classifiers are formalized in
\texttt{sub\_object\_classifier.v}.% (284 lines).
%
\item Colimits and \v{C}ech nerves are formalized in
\texttt{Limit, T.v, OT.v OT\_Tf.v, T\_telescope.v, Tf\_Omono\_sep.v}.% (552 lines).
\item
Reflective subuniverses and modalities are formalized in
\texttt{reflective\_subuniverse.v, modalities.v}. % (1053 lines).
\item 
%
  The definition of the dense topology as a left exact modality on
  $\HProp$ is given in \texttt{sheaf\_base\_case.v}. % (186 lines).
\item
Section~\ref{sec:sheaves} is formalized in
\texttt{sheaf\_def\_and\_thm.v}. % (1029 lines).
\item
Section~\ref{sec:sheafification} is formalized in
\texttt{sheaf\_induction.v}. % (2340 lines).
\end{itemize}

@@

In the formalization, we had to use the type-in-type option, to handle
the universe issues we faced.

@@

%
If Coq handles cumulativity on $\Type$ natively, it is not
the case for the $\Sigma$-type $\Type_n$, which require propositional
resizing. 
%
This issue could be solved by adding an axiom of cumulativity
for $\Type_n$ with an explicit management of universes. 
%
But as it would not have any computational content, such a solution
would really complicate the proofs as the axiom would appear
everywhere cumulativity is needed and it would need explicit
annotations for universe levels everywhere in the formalization.
%
When the absence of cumulativity on $\Type_n$ rises a universe issue
for a lemma (it occurs twice in the formalization), we have proved the
lemma and then admit the exact same lemma, just letting Coq infer less
constrained universe levels.

@@

As a future work, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  At the moment, sheafification only works with truncated
  types. This is fine from a programming point of view, 
  but many important types in homotopy theory are not
  truncated, \eg the circle $\mathbb S^2$.  Thus, we plan to extend
  the sheafification to all types by using a
  limit construction.
\item 
  We would like to have much more examples of left-exact modalities on
  $\HProp$, in order to extend type theory with different principles
  than excluded middle.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Grothendieck~\cite[Section~V.4]{maclanemoerdijk} in the sense that any
Grothendieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.

% conference papers do not normally have an appendix

% use section* for acknowledgment
\section*{Acknowledgment}

@@

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
\bibliography{sheaf_lics}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
% \begin{thebibliography}{5}

% that's all folks

>>>>

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint,10pt,numbers]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

\usepackage{amsmath}

@@

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
\usepackage{amssymb,amsthm,amsfonts}

\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{defi}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{ax}[thm]{Axiom}
\newtheorem{cond}[thm]{Condition}
\newtheorem{exm}[thm]{Example}

@@

\newcommand{\cL}{{\cal L}}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{LICS '16}{July 5--8, 2016, New York City, USA}
\copyrightyear{2016}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Lawvere-Tierney Sheafification\\ in Homotopy Type Theory}
% \subtitle{Subtitle Text, if any}

\authorinfo{Kevin Quirin}
           {\'Ecole des Mines de Nantes}
           {kevin.quirin@inria.fr}
\authorinfo{Nicolas Tabareau}
           {Inria}
           {nicolas.tabareau@inria.fr}

\maketitle

\begin{abstract}
  Sheafification is a popular tool in topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been developed in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher toposes.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, being
  compatible with the full type-theoretic axiom of choice.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

@@

One of the most direct application of homotopical notion to type
theory is the introduction of homotopy types. 
%
Using the analogy that points in a space correspond to elements of a
type and that paths between two points correspond to 
elements of the corresponding identity type (which defines equality in type theory),  
%
an $n$-type is simply a type
for which equality becomes trivial up to level $n$. 
%
Voevodsky has realized that this notion admits a compact inductive definition
internal to type theory, given by
% Our work is mainly based on the stratification by $\Type_n$~:
\begin{defi}
  $\IsType n$ is defined by induction on $n\geqslant -2$:
  \begin{itemize}
  \item $\IsType {(-2)} X$ if $X$ is a contractible type, \ie $X$
    is pointed by $c:X$, and every other point in $X$ is connected to $c$.
  \item $\IsType {(n+1)} X = \prod_{x,y:X} \IsType n (x=y).$
  \end{itemize}
  Then, $\Type_n = \sum_{X:\Type} \IsType n X$.
\end{defi}
% For a type $X$, to be in $\Type_n$ means that paths spaces of $X$ are
% trivial after $n+1$ iteration.
%
We also define some syntactic sugar for contractible types and mere propositions.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
$$
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
$$ 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
\begin{align*}
  \chi &: \left(\sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b}\right) \\ &\qquad\xrightarrow{\sim} 
 (B \to \Type_n)
\end{align*}
%
 such that the diagram
$$
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
$$
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
$$t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).$$

@@

Then $P$ is the {\em colimit} of $D$ if for any other $X:\Type$, the
map
$$\lambda f:P \to X, \left( \lambda i,~f \circ q_i\, ;\, \lambda i\, j\,
  \phi,\ \precompose_f (p_{i, j}^\phi)) \right)$$
where $\precompose_f : \phi = \psi \to \phi \circ f = \psi \circ f$,
is an equivalence.
\end{defi}

@@

\begin{lem}
  Let $D$ be a diagram, $m$ a truncation index, and
  $P:\Type_{m}$. Let's call $P$ a $m$-colimit of $D$ if
  for any other $X:\Type_m$, the
  map
  $$\lambda f:P \to X, \left( \lambda i,~f \circ q_i\, ;\, \lambda i\, j\,
  \phi,\ \precompose_f (p_{i, j}^\phi)) \right)$$
  is an equivalence.

@@

\begin{defi}
  Let $f:X \to Y$ be a map. The kernel pair $T_f$ of $f$ is the higher inductive type given
  by
  \begin{itemize}
  \item $t:~X \to T_f$
  \item $\alpha:~\forall a\,b:X,~f a = f b \to t a = t b$
  \item $\alpha_1:~\forall a:X,~\alpha\, a\, a\, 1 = 1$
  \end{itemize}
  We view $T_f$ as the coequalizer of
  $$\xymatrix{
    \sum_{a,b:X} f a = f b \dar[r]{4pt}^-{\pi_1}_-{\pi_2} & X
  }$$
  preserving the identity.

@@

Then, the desired diagram is the mapping telescope of the iterations
of $T$.
\begin{defi}
  Let $f$ be a map from $X$ to $Y$. Then the iterated kernel pair of
  $f$ $C(f)$
  is given by the diagram
  $$\xymatrix{
    C(f) := X \ar[r]^t & T_f \ar[r]^t & T_{\tilde f} \ar[r] &\cdots
  }$$
\end{defi}

@@

The proof is based on the following central result.
\begin{lem}
  Let $X:\Type_p$, $Y:\Type_p^\modal$ and $f:X\to Y$. If for all $y:Y$,
  $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
\end{lem}
%
It is straighforward to define a map
$$\phi:\sum_{x:X}  (f x = y)\to
\sum_{x:\modal X} (\modal f x = \eta_Y y),$$
using $\eta$ functions.
We just need to check that every $\modal$-fiber $\modal(\fib \phi {x;p})$ is
contractible.
Technical transformations allow one to prove
$$\fib\phi{x;p} = \fib s {y;p^{-1}}$$
for some $a:\modal X$, $b:\modal Y$ and 
$$s:\fib{\eta_X}a \to \fib{\eta_Y}b.$$
%
From definition of left exactness, one can deduce the following:
\begin{lem}
  Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
  contractible, then so is $\fib f b$ for any $b:B$.
\end{lem}
Thus, we just need to prove that $\modal(\fib {\eta_X} a)$ and
$\modal(\fib {\eta_Y} b)$ are contractible.
But we have the lemma:
\begin{lem}
  Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
  y)$ is contractible.
\end{lem}

@@

Here, we need to distinguish between
dense $(-1)$-sub\-objects, that will be used in the definition of
sheaves, and dense $n$-subobjects, that will be used in the definition
of separated types.

@@

  We define $g$ as 
  $$g(e) = \modal_n \left( \fib \phi e \right), $$
  where
  $$\phi : \left(\sum_{b:\left(\sum_{e:E} (\chi\, e)\right)} (f\,
    b)\right) \to E$$
  defined by $\phi(x) = (x_1)_1$.
  Using the following properties on fibers, we can prove that it
  defines an equivalence.
  \begin{lem}
    Let $A,B,C:\Type_n$, $f:A\to B$ and $g:B\to C$.
    Then
    \begin{itemize}
    \item If $g$ is an embedding, then 
      $$\prod_{b:B}\fib f b = \fib {g\circ f} {g\, b}.$$
    \item If all fibers of $f$ and $g$ are $n$-truncated, then
      $$\prod_{c:C}\modal_n(\fib {g \circ f} c) =
      \modal_n \left(  
        \sum_{w:\fib g c} \modal_n (\fib f {w_1})
      \right).$$
    \end{itemize}
  \end{lem}

@@

Let $T : \Type_{n+1}$. We define $\separated T$ as the image (see
Section~\ref{sec:epi-mono-fact}) of
$\modal_n^T \circ \{\cdot\}_T$, as in
$$\xymatrix{
  T \ar[r]^{\{\cdot\}_T} \ar[d]_{\mu_T} & \Type_n^T \ar[d]^{\modal_n^T} \\
  \separated T \ar[r]& \left( \Type_n^\modal \right)^T
}, $$
where $\{\cdot\}_T$ is the singleton map $\lambda (t:T),~\lambda
(t':T),~t=t'$.
%
$\separated T$ can be given explicitly by
%
$$
\begin{array}{rcl}
\separated T &\defeq & \im (\lambda~t:T,~\lambda~ t',~ \modal_n (t = t')) \\
          & \defeq & \sum\limits_{u:T \to \Type_n^\modal} \left\| \sum_{a:T} 
            (\lambda t,~\modal_n (a=t)) = u\right\|.
\end{array}
$$
%
This corresponds to the free separated object used in the topos-theoretic construction, but using $\Type_n^\modal$ instead of the
$j$-subobject classifier $\Omega_j$.
%
\begin{prop}
  For any $T:\Type_{n+1}$, $\separated T$ is separated.  
\end{prop}

@@

  We use the encode-decode~\cite[Section 8.9]{hottbook} method to characterize $\tilde f |a|_{n+1}
  = x$, and the result follows. We refer to the formalization for details.
\end{proof}

@@

\section{Extension to Type}
\label{sec:extension-type}

In the previous section, we defined a (countably) infinite family of
modalities $\Type_i \to \Type_i$. One can extend them to whole
$\Type$ by composing with truncation:

\begin{lem}\label{lem:type}
  Let $\modal_i:\Type_i \to \Type_i$ be a modality. Then $\modal
  \defeq \modal_i
  \circ \|\cdot\|_i : \Type \to \Type$ is a modality in the sense
  of~\cite[Section 7.7]{hottbook}
\end{lem}

\begin{exm}
  If $\modal_{-1}$ is the double negation modality on $\HProp$ and
  $i=-1$, for any type $X$:
  \begin{align*}
  \modal X &\defeq \lnot\lnot \|X\|  \\
           &= \lnot\lnot X
  \end{align*}
  by univalence for h-propositions.

 Thus, $\modal$ is exactly the double negation modality on $\Type$
  described in~\ref{sec:examples-left-exact}.
\end{exm}

Chosing $i\geqslant 0$ is a refinement of this double negation
modality on $\Type$: it will collapse every type to a $\Type_i$,
instead of an $\HProp$.

Taking $i$ as a non-fixed parameter, that would allow us to work in an
universe where the new principle ({\em e.g.} mere excluded middle) is
true for any explicit truncated type. Indeed, $i$ can be chosen
dynamically along a proof, and thus be increased as much as needed.

Unfortunately, one lose canonicity when working in $\Type^\modal$;
indeed, for example, the term 
$$\left( \lambda P:\HProp,~\left\{\begin{array}{l} 0 \text{ if $P$ is
                                    false} \\ 1 \text{ if $P$ is
                                    true} \end{array}\right. \right)
                            P~:~\nat$$
will not reduce to a numeral if the chosen property $P$ is not
decidable. However, we keep canonicity for ``old'' terms, i.e. the
ones constructed without using excluded middle.

% \subsection{New objects}
% \label{sec:new-objects}

% Working in $\Type^\modal$, one can consider new objects. For example,
% let $c:nat \to \nat$ implementing the Collatz function. Then, for any
% $n:\nat$, the type $$C_n \defeq \sum_{k:\nat}\left( c^k(n) = 1 \land k\text{ is the
%   smallest such $\nat$}\right)$$ is an $\HProp$.
% Thus, for any $n$, one have $C_n + \lnot C_n$, and one can {\em
%   constructively} build a map $\varphi:\nat \to \two$ such that
% $\varphi(n) = \True$ if and only if $n$ satisfies the Collatz
% conjecture.
% Actually, one can defined such a map for any family of $\HProp$
% indexed by $\nat$, for example with the family $C_n \defeq \text{«\,The $n$-th Turing Machine halts\,»}$.

@@

A Coq formalization based on the Coq/HoTT library~\cite{hottlib} is available at
\url{github.com/KevinQuirin/sheafification}.
%
After reviewing the content and some statistics about the
formalization in Section~\ref{sec:cont-form}, we present the
limitations of our formalization in Section~\ref{sec:limit-form}, in
particular the issues relative to universe polymorphism.

@@

We provide a more detailed insight of the structure of our formalization:
\begin{itemize} 
\item Colimits and iterated kernel pairs are formalized in
\texttt{Limit}, \texttt{T.v}, \texttt{OT.v}v \texttt{OT\_Tf.v}, \texttt{T\_telescope.v},\\ \texttt{Tf\_Omono\_sep.v}.% (552 lines).
\item
Reflective subuniverses and modalities are formalized in\\
\texttt{reflective\_subuniverse.v}, \texttt{modalities.v}. % (1053 lines).
\item 
%
  The definition of the dense topology as a left exact modality on
  $\HProp$ is given in \texttt{sheaf\_base\_case.v}. % (186 lines).
\item
Section~\ref{sec:sheaves} is formalized in
\texttt{sheaf\_def\_and\_thm.v}. % (1029 lines).
\item
Section~\ref{sec:sheafification} is formalized in
\texttt{sheaf\_induction.v}. % (2340 lines).
\end{itemize}

@@

In the formalization, we had to use the \texttt{type-in-type} option, to handle
the universe issues we faced.

@@

%
If Coq handles cumulativity on $\Type$ natively, it is not
the case for the $\Sigma$-type $\Type_n$, which require propositional
resizing. 
%
This issue could be solved by adding an axiom of cumulativity
for $\Type_n$ with an explicit management of universes. 
%
But as it would not have any computational content, such a solution
would really complicate the proofs as the axiom would appear
everywhere cumulativity is needed and it would need explicit
annotations for universe levels everywhere in the formalization.
%
Note that we also could have resolved the issue by giving explicitely
the universe levels we wanted, like in the Coq/HoTT library\cite{hottlib}.

@@

As a future work, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  The extension to whole $\Type$ in lemma~\ref{lem:type} is not
  totally satisfactory, as every type is collapsed to a truncated
  one. But some types in homotopy type theorey are not
  truncated\cite[Example 8.8.6]{hottbook}.

  % At the moment, sheafification only works with truncated
  % types. This is fine from a programming point of view, 
  % but many important types in homotopy theory are not
  % truncated, \eg the circle $\mathbb S^2$.  Thus, we plan to extend
  % the sheafification to all types by using a
  % limit construction.
\item 
  We would like to have much more examples of left-exact modalities on
  $\HProp$, in order to extend type theory with different principles
  than excluded middle.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Gro\-thendieck~\cite[Section~V.4]{maclanemoerdijk} in the sense that any
Grothen\-dieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.

\section*{Acknowledgment}

@@

\bibliographystyle{abbrvnat}
\bibliography{sheaf_lics}
